\documentclass[a4paper,10pt]{article}
% Use ctrl + alt + V to view live pdf

% Packages
\usepackage[utf8]{inputenc} % For encoding
\usepackage[T1]{fontenc} % Better handling of accented characters and hyphenation
\usepackage{microtype} % Improves spacing and justification
\usepackage{amsmath, amssymb} % For equations and symbols
\usepackage{graphicx} % For including graphics/images
\usepackage{caption} % For customizing figure and table captions
\usepackage{subcaption} % For subfigures and subcaptions
\usepackage{float} % For fixing figure and table positions
\usepackage{booktabs} % For professional-looking tables
\usepackage{siunitx} % For consistent typesetting of units and numbers
\usepackage[margin=2cm]{geometry} % Adjusts page margins
\usepackage{fancyhdr} % For custom headers and footers
\usepackage{lmodern} % For a professional-looking font (main body font)
\usepackage{titlesec} % For title customization
\usepackage{array} % For custom table formatting
\usepackage[colorlinks=true, linkcolor=black, urlcolor=black]{hyperref} % Colored links without boxes
\usepackage{cleveref} % For improved cross-referencing    
\usepackage{multirow}
\usepackage{enumitem}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{textcomp}
\usepackage{tabularx}
\usepackage{changepage}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows}
\newcolumntype{Y}{>{\centering\arraybackslash}X}


\lstdefinestyle{vhdl-style}{
    language=VHDL,
    basicstyle=\ttfamily\footnotesize,
    keywordstyle=\bfseries\color{blue},
    commentstyle=\itshape\color{gray},
    stringstyle=\color{red},
    numbers=left,
    numberstyle=\tiny\color{gray},
    stepnumber=1,
    breaklines=true,
    showstringspaces=false,
    frame=single
}
\lstset{style=vhdl-style}
\lstset{captionpos=b}
\lstset{basicstyle=\ttfamily\scriptsize} 
\renewcommand{\lstlistingname}{Program}

% Custom settings
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textit{GB3 - Risc-V Processor}} % Header left
\fancyhead[R]{\textit{Will Hewes - wh365}} % Header right 
\fancyfoot[C]{\thepage} % Footer center
\setlength{\headheight}{15pt} % Header height
\setlength{\parindent}{0em} % Indentation for paragraphs
\setlength{\parskip}{0.2em} % Add spacing between paragraphs
\setlength{\abovedisplayskip}{0.5em}
\setlength{\belowdisplayskip}{0.5em}
\setlength{\abovedisplayshortskip}{0.5em}
\setlength{\belowdisplayshortskip}{0.5em}
% \setlist{topsep=0em, partopsep=0em, itemsep=0em, parsep=0em}

\graphicspath{{Images/}}

% \renewcommand{\arraystretch}{1.2}

% Title formatting
\renewcommand{\maketitle}{
    \begin{center}
        \LARGE \textbf{ENGINEERING TRIPOS PART IIA} \\ 
        \vspace{0.5em}
        \Large \textbf{GB3 - Risc-V Processor} \\ 
        \vspace{0.5em}
        \textbf{First Interim Report} \\
        \large Group 4 - Resource Usage \\
        \vspace{1em}
        \large Will Hewes - wh365 \\ 
        Pembroke College \\ 
        \vspace{0.5em}
    \end{center}
}

\begin{document}
\maketitle
\hrule
\tableofcontents
\newpage

\section{Introduction}
\label{sec:Introduction}
% This section should give a quick overview of the team’s task as a whole

This project involves the collaborative design and implementation 
of a RISC-V processor on an iCE40 FPGA device. 
The overarching aim is to optimise 
the processor design with respect to its 
performance, power dissipation, and resource usage, 
while balancing trade-offs between these three aspects. 
Each team member is responsible for focusing on one of the three aspects.
My reports will focus on the resource usage of the design.

This interim report characterises the baseline design. 
We are evaluating three provided example programs - 
\texttt{hardwareblink}, \texttt{softwareblink}, and \texttt{bubblesort} - 
to establish benchmarks for the logic usage, power dissipation and timing. 
These programs utilise the processor and surrounding hardware to varying degrees, 
and provide a baseline against which future design improvements can be compared.


\section{Project Specification}
\label{sec:Project_Specification}
%This section should provide an overview of your specific
%task as part of the team.

Within the team, my specific task focuses on 
optimising the resource usage of the FPGA implementation. 
The goal of this optimisation is to reduce the number of 
Look-Up Tables (LUTs), flip-flops, and logic elements used in the design. 
This allows more headroom on the device for future expansions 
and increases its scalability.

LUTs and flip-flops are some of
the fundamental building blocks of digital logic in an FPGA. 
A LUT is a configurable logic element that can implement 
arbitrary combinational logic functions 
by storing precomputed outputs for each possible input combination. 
Flip-flops are used to store single bits of state and 
are essential for implementing sequential logic, 
such as registers and control signals. 
The quantity and efficiency of these elements directly impact 
how much logic the FPGA can implement and how complex the system can be.

Optimising resource usage means reducing 
how many of these elements a design uses.
By simplifying logic paths, reusing hardware blocks,
or packing multiple functions into fewer LUTs and flip-flops,
space can be freed up on the FPGA.
This can come at the cost of slower computational speeds
or increased power dissipation depending on how these changes are implemented.
This represents the crux of the project,
and future reports will aim to balance these constraints against each other.

For this interim report, the goal is not to 
change the programs or optimise the device yet.
This report only characterises the resource usage of the three programs,
providing a basis for analysing resource usage in future designs.

\section{Summary of Preliminary Design Work}
\label{sec:Summary_of_Preliminary_Design_Work}

To assess the resource usage of each of the three programs, 
the designs were synthesised using \texttt{yosys} 
and place-and-route was performed using \texttt{nextpnr}. 

The resulting resource usage reports of these processes 
are displayed in the Appendix.

\subsection{Hardwareblink}
\label{sec:Hardwareblink}

The first program evaluated was \texttt{hardwareblink}.
This example does not use the RISC-V processor core and 
instead relies purely on combinational and 
sequential logic defined in \texttt{Verilog}.

The program implements a simple hardware-only design that 
toggles \textit{LED D14} at 1\,Hz using a counter. 
In order to toggle the LED at 1\,Hz, 
\texttt{hardwareblink} makes use of the 48\,MHz \textit{HFOSC},
and divides it by 24\,000\,000 in a counter.
This output toggles a register, 
the value of which is mapped to pin \textit{D3} (driving \textit{LED D14}),
giving rise to a 1\,Hz blinking light.
The summary of the resulting logic utilisation is shown
in Table \ref{tab:hardware_yosys_report} in the Appendix.

The design uses 33 flip-flops,
61 carry-chain elements,
65 combinational logic cells,
the high frequency clock,
and no block RAMs.
This corresponds to 160 cells in total.

Subsequently, \texttt{nextpnr} was used to place and route these logic cells,
with the results shown in Table \ref{tab:hardware_pnr_report} in the Appendix.
This report shows that only 103 logic cells were used in this implementation, 
corresponding to approximately 2\% of the available 5280 logic cells 
on the device.

No block RAMs, memories, or processor logic were involved in this design, 
making it a minimal baseline for comparison. 
The low utilisation confirms the simplicity of the circuit 
and provides a reference point for understanding 
how resource usage increases with more complex processor-driven programs.

\subsection{Softwareblink}
\label{sec:Softwareblink}

The second program evaluated was \texttt{softwareblink}. 
In contrast to \texttt{hardwareblink}, 
this design makes use of the RISC-V processor core 
to execute a compiled C program that toggles \textit{LED D14} via software. 
The program operates by writing to a memory-mapped I/O register at regular intervals, 
with timing determined by a simple delay loop. 
This introduces a significant increase in logic complexity, 
as instruction decoding, arithmetic, control logic, and memory access logic 
are all active.

Table \ref{tab:software_yosys_report} in the Appendix 
shows the breakdown of primitive cell usage 
as reported by \texttt{yosys} after synthesis.

The design uses 673 flip-flops (and three not shown in the report),
61 carry-chain elements,
1965 combinational logic cells,
the high frequency clock,
and 20 block RAMs.
This corresponds to 2878 cells in total.

Evidently this software design uses far more logic elements
than the previous hardware design,
requiring a significant portion of the total RAM capacity of the device.

Table \ref{tab:software_pnr_report} in the Appendix
shows the corresponding logic usage 
on the FPGA as reported by \texttt{nextpnr}. 
This design takes up 50\% of the device's logic cell resources 
and two-thirds of its available block RAMs, 
representing a substantial increase in resource usage 
relative to \texttt{hardwareblink}.

This evaluation establishes the baseline resource cost 
of running programs on the processor. 
The increased logic and memory usage, relative to \texttt{hardwareblink}, 
reflect the added complexity of 
instruction decoding, pipeline control, and memory access. 
These results provide a benchmark for future optimisations 
aimed at reducing the resource footprint of the processor.

\subsection{Bubblesort}
\label{sec:Bubblesort}

The third program evaluated was \texttt{bubblesort}, 
which tests the processor under a significantly 
heavier computational load compared to \texttt{softwareblink}. 
The program implements a nested loop to sort a small array in memory, 
exercising the arithmetic, comparison, branching, 
and memory access capabilities of the RISC-V core. 
This makes it a more demanding benchmark for 
evaluating resource utilisation.

Table \ref{tab:bubblesort_yosys_report} in the Appendix
shows the resulting breakdown of 
primitive cells as reported by \texttt{yosys},
and the corresponding place-and-route results reported by 
\texttt{nextpnr} are shown in 
Table \ref{tab:bubblesort_pnr_report} in the Appendix.

This design required a larger number of LUTs than 
\texttt{softwareblink}, increasing from 1\,965 to 2\,344, 
as the processor executes a more instruction-heavy and 
branching-intensive routine. 
However, the number of flip-flops, RAM blocks, 
and carry-chain elements remained constant. 
This suggests that the core hardware logic is unchanged, 
and the increase in combinational logic reflects 
the execution of a more complex program 
rather than architectural modifications.

\section{Future Work}
\label{sec:Future_Work}

Looking at the \texttt{yosys} and \texttt{nextpnr} reports 
for the three baseline programs, it is clear that the primary constraints 
placed on resource usage occur due to software complexity, 
particularly in terms of block RAM and logic cell utilisation.

In order to optimise resource usage, 
the focus will be on reducing logic complexity introduced by 
processor control and data path operations. 
This may include combining pipeline registers where possible,
removing support for unused instructions, 
and relocating the general-purpose register file into a single block RAM
to reduce flip-flop usage.
Where feasible, simpler control schemes that mirror the low-overhead structure of 
\texttt{hardwareblink} will be adopted. 
These changes should reduce the number of LUTs required, 
primarily by reducing the amount of combinational logic needed for control flow, 
arithmetic, and memory interfacing.

\section{Conclusion}
\label{sec:Conclusion}

A baseline characteristic for the resource usage has been determined.
Looking at the outputs of the synthesis and place-and-route reports,
the primary reason for greater resource usage comes down to
the software complexity applied to the FPGA.

The \texttt{hardwareblink} program had the lowest resource usage,
relying on no block RAM and
only taking up roughly 2\% of the available logic cells.
It achieved this by avoiding the use of the RISC-V processor entirely, 
instead implementing the desired functionality directly in Verilog 
as a simple counter and output toggle. 

In contrast, both \texttt{softwareblink} and \texttt{bubblesort} 
introduced significantly more complexity due to 
the demands of instruction execution, 
memory access, and control logic. 
These additional demands used over 50\% of the available logic cells 
and took up two thirds of the available block RAM.

In order to minimise the resource usage associated with FPGA implementations,
future work will focus on optimising the processor microarchitecture.
This includes combining pipeline registers where feasible, 
and cutting down on unnecessary logic functions.
Each change will be weighted against 
the performance and power dissipation constraints.

\newpage
\appendix
\label{Appendix}
\section{Resource Usage Data}
%Use this section to include diagrams, Verilog or C code, etc

\begin{table}[H]
    \centering
    \begin{tabular}{|l|c|}
        \hline
        \textbf{Primitive} & \textbf{Count} \\
        \hline
        \texttt{SB\_LUT4} logic cells & 65 \\
        \texttt{SB\_DFF} flip-flops & 0 \\
        \texttt{SB\_DFFE} flip-flops with clock enable & 1 \\
        \texttt{SB\_DFFSR} flip-flops with set/reset & 32 \\
        \texttt{SB\_DFFSS} flip-flops with set/set & 0 \\
        \texttt{SB\_CARRY} carry-chain elements & 61 \\
        \texttt{SB\_RAM40\_4K} block RAMs & 0 \\
        \texttt{SB\_HFOSC} internal oscillator & 1 \\
        \hline
    \end{tabular}
    \caption{Primitive cell usage reported by 
    \texttt{yosys} for the \texttt{hardwareblink} design}
    \label{tab:hardware_yosys_report}
\end{table}

\begin{table}[H]
    \centering
    \begin{tabularx}{0.6\textwidth}{X c c}
        \toprule
        Resource type & Used & \% of total \\ \midrule
        Logic cells (\texttt{ICESTORM\_LC}) & 103 / 5\,280 & 1.9\,\% \\
        Block RAMs (\texttt{ICESTORM\_RAM}) & 0 / 30 & 0\,\% \\
        IO buffers (\texttt{SB\_IO}) & 1 / 96 & 1.0\,\% \\
        Global buffers (\texttt{SB\_GB}) & 2 / 8  & 25\,\% \\
        HF oscillators (\texttt{SB\_HFOSC}) & 1 / 1  & 100\,\% \\ 
    \bottomrule
    \end{tabularx}
    \caption{Place-and-route logic utilisation reported by 
    \texttt{nextpnr} for \texttt{hardwareblink}}
    \label{tab:hardware_pnr_report}
\end{table}

\begin{table}[H]
    \centering
    \begin{tabular}{|l|c|}
        \hline
        \textbf{Primitive} & \textbf{Count} \\
        \hline
        \texttt{SB\_LUT4} logic cells & 1\,965 \\
        \texttt{SB\_DFF} flip-flops & 415 \\
        \texttt{SB\_DFFE} flip-flops with clock enable & 154 \\
        \texttt{SB\_DFFSR} flip-flops with set/reset & 100 \\
        \texttt{SB\_DFFSS} flip-flops with set/set & 4 \\
        \texttt{SB\_CARRY} carry-chain elements & 216 \\
        \texttt{SB\_RAM40\_4K} block RAMs & 20 \\
        \texttt{SB\_HFOSC} internal oscillator & 1 \\
        \hline
    \end{tabular}
    \caption{Primitive cell usage reported by 
    \texttt{yosys} for the \texttt{softwareblink} design}
    \label{tab:software_yosys_report}
\end{table}

\begin{table}[H]
    \centering
    \begin{tabularx}{0.6\textwidth}{X c c}
        \toprule
        Resource type & Used & \% of total \\ \midrule
        Logic cells (\texttt{ICESTORM\_LC}) & 2662 / 5280 & 50\,\% \\
        Block RAMs (\texttt{ICESTORM\_RAM})& 20 / 30 & 66\,\% \\
        IO buffers (\texttt{SB\_IO}) & 8 / 96 & 8\,\% \\
        Global buffers (\texttt{SB\_GB}) & 5 / 8 & 62\,\% \\
        HF oscillators (\texttt{ICESTORM\_HFOSC}) & 1 / 1 & 100\,\% \\
        \bottomrule
    \end{tabularx}
    \caption{Place-and-route logic utilisation reported by 
    \texttt{nextpnr} for \texttt{softwareblink}}
    \label{tab:software_pnr_report}
\end{table}

\begin{table}[H]
    \centering
    \begin{tabular}{|l|c|}
        \hline
        \textbf{Primitive} & \textbf{Count} \\
        \hline
        \texttt{SB\_LUT4} logic cells & 2\,344 \\
        \texttt{SB\_DFF} flip-flops & 415 \\
        \texttt{SB\_DFFE} flip-flops with clock enable & 154 \\
        \texttt{SB\_DFFSR} flip-flops with set/reset & 100 \\
        \texttt{SB\_DFFSS} flip-flops with set/set & 4 \\
        \texttt{SB\_CARRY} carry-chain elements & 216 \\
        \texttt{SB\_RAM40\_4K} block RAMs & 20 \\
        \texttt{SB\_HFOSC} internal oscillator & 1 \\
        \hline
    \end{tabular}
    \caption{Primitive cell usage reported by 
    \texttt{yosys} for the \texttt{bubblesort} design}
    \label{tab:bubblesort_yosys_report}
\end{table}

\begin{table}[H]
    \centering
    \begin{tabularx}{0.6\textwidth}{X c c}
        \toprule
        Resource type & Used & \% of total \\ \midrule
        Logic cells (\texttt{ICESTORM\_LC}) & 3073 / 5280 & 58\,\% \\
        Block RAMs (\texttt{ICESTORM\_RAM}) & 20 / 30 & 66\,\% \\
        IO buffers (\texttt{SB\_IO}) & 8 / 96 & 8\,\% \\
        Global buffers (\texttt{SB\_GB}) & 5 / 8 & 62\,\% \\
        HF oscillators (\texttt{ICESTORM\_HFOSC}) & 1 / 1 & 100\,\% \\
        \bottomrule
    \end{tabularx}
    \caption{Place-and-route logic utilisation reported by 
    \texttt{nextpnr} for \texttt{bubblesort}}
    \label{tab:bubblesort_pnr_report}
\end{table}

\end{document}